Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-06-03 ms.tex
Nearing final draft of intro paragraph

The outer layers of planetary atmospheres are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation and regulate the escape of gas into space. In atmospheres with thick layers of atomic hydrogen, stellar Lyman Alpha (Ly$\alpha$) radiation is especially important as the optical depth at line center is large. Ionization and subsequent recombination can generate \lya within the planet's atmosphere, dissociating molecules in lower layers and causing heating by collisional de-excitation. It can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy.

@@

\ifx
\todo{
Hubble Space Telescope observations have found large Lyman $\alpha$ transit depths around the gas giants HD 209458b \citep{2003Natur.422..143V}, HD 189733b \citep{2012A&A...543L...4L} and 55 Cnc b \citep{2012A&A...547A..18E}, Neptune-size planets GJ 436 b 
\citep{2015Natur.522..459E}
%, 2017A&A...605L...7L,2019A&A...629A..47D}   
and GJ 3470 b \citep{2018A&A...620A.147B}, and possibly the super-Earth HD 219134b \citep{2019EPSC...13.1928L}.
Non-detections or marginal detections have been reported for the super-Earths 55 Cnc e \citep{2012A&A...547A..18E}, super-Earth HD 97658 b \citep{2017A&A...597A..26B} , super-Earth GJ 1132 b \citep{2019AJ....158...50W}, super-Earth $\pi$ Men c \citep{2020ApJ...888L..21G}, and marginal detections were  Earth-size TRAPPIST-1 \citep{2017A&A...599L...3B} and sub-Earth Kepler-444 \citep{2017A&A...602A.106B}.
}
\fi

@@

Hubble Space Telescope (HST) observations with STIS have found large Lyman $\alpha$ transit depths around the gas giants HD 209458b \citep{2003Natur.422..143V}, HD 189733b \citep{2012A&A...543L...4L} and 55 Cnc b \citep{2012A&A...547A..18E}, Neptune-size planets GJ 436 b
\citep{2015Natur.522..459E}
and GJ 3470 b \citep{2018A&A...620A.147B}, and possibly the super-Earth HD 219134b \citep{2019EPSC...13.1928L}. In the case of GJ 436 b, ultraviolet transit depths of 56\% are reported, claiming a total Ly-$\alpha$ eclipse would have been possible for a central transit. Follow-up observations of GJ 3470 b by \citet{2021arXiv210309864B} confirm the presence of an exospheric hydrogen cloud around the planet with a radius of 1.3 times its Roche lobe in the far-ultraviolet, implying that H atoms are escaping the atmosphere. 
Non-detections or marginal detections have been reported for the super-Earths 55 Cnc e \citep{2012A&A...547A..18E}, super-Earth HD 97658 b \citep{2017A&A...597A..26B} , super-Earth GJ 1132 b \citep{2019AJ....158...50W}, super-Earth $\pi$ Men c \citep{2020ApJ...888L..21G}, and marginal detections were Earth-size TRAPPIST-1 \citep{2017A&A...599L...3B} and sub-Earth Kepler-444 \citep{2017A&A...602A.106B}.  \cite{2021ApJ...907L..36M} reports a detection of CII ions in a single transit of the exoplanet $\pi$ Men c, with transit depths and Doppler velocities consistent with an escaping atmosphere.

These observations have revealed a population of atoms extending out to distances of order the planetary radius for several planets around bright, nearby stars. The importance of the \lya line in the atmospheric dynamics of these systems necessitates a careful treatment of resonance-line radiation transfer in order to construct accurate models of atmospheric heating and escape. However, the extreme optical depths at \lya line center introduce a computational barrier to simulating radiative transfer via Monte Carlo methods.

The main impediment to peforming magnetohydrodynamic simulations with fully a coupled radiative transfer solution via Monte Carlo methods is the prohibitive computational cost. A primary factor in the cost of Monte Carlo is that the number of scatterings a photon (packet/superphoton?) undergoes is proportional to the line center optical depth.  Near the base of the atmosphere individual simulation cells can have optical depths of more than a million so most of the time is spent integrating photons in these cells. This means that any method that can treat these zone via an analytic solution has the potential to greatly accelerate the calculation.

There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations.  The simplest class of methods that is specific to resonance line transfer are core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A}.  These methods focus Monte Carlo transfer on the wings of the line and avoid scatterings in the core. Line core photons are pushed to the wings by biasing the atomic velocity distribution to high velocities so tha opacity is lower and escape is more rapid.

A second class of methods are hydbrid diffusion methods \citep{2001JCoPh.172..543G}, such as discrete diffusion Monte Carlo \citep{2007JCoPh.222..485D}. This involves designating sufficiently optically thick simulations cells for computation via a finite difference representaiton of radiation transfer equation. This method is particularly attractive for gray opacity applications where a single diffusion equation can be computed.  For the resonant scattering problem, where the opacity is a strong function of frequnecy, implementation of this method is more complex and computationally demanding as it requires the solution of many frequency groups, but can lead to a considerable speed up \citep{2018MNRAS.479.2065S}.

@@

>>>>

The outer layers of planetary atmospheres are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation and regulate the escape of gas into space. In atmospheres with thick layers of atomic hydrogen, stellar Lyman Alpha (Ly$\alpha$) radiation is especially important in this process as the optical depth at line center is large. Ionization and subsequent recombination can generate \lya within the planet's atmosphere, dissociating molecules in lower layers and causing heating by collisional de-excitation. It can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy.

@@

@@

Hubble Space Telescope (HST) observations with STIS have found large Lyman $\alpha$ transit depths, indicating an extended distribution of hydrogen around a handful of exoplanets \todo{Organize these by date?} (\citealp{2003Natur.422..143V}; \citealp{2012A&A...543L...4L}; \citealp{2012A&A...547A..18E}; \citealp{2015Natur.522..459E}; \citealp{2018A&A...620A.147B}; \citealp{2019EPSC...13.1928L}; \citealp{2021arXiv210309864B}; \citealp{2012A&A...547A..18E}; \citealp{2017A&A...597A..26B}; \citealp{2019AJ....158...50W}; \citealp{2020ApJ...888L..21G}; \citealp{2017A&A...599L...3B}; \citealp{2017A&A...602A.106B}). These observations have revealed a population of atoms extending out to distances of order the planetary radius for several planets around bright, nearby stars, motivating a study of the physics of \lya interactions with H atoms surrounding a planet. The importance of the \lya line in the atmospheric dynamics of these systems necessitates a careful treatment of resonant scattering in order to construct accurate models of atmospheric heating and escape. However, the extreme optical depths at \lya line center introduce a computational barrier to simulating radiative transfer via Monte Carlo methods.

The main impediment to performing magnetohydrodynamic simulations with fully a coupled radiative transfer solution via Monte Carlo methods is the prohibitive computational cost. A primary factor in the cost of Monte Carlo is that the number of scatterings a photon undergoes is proportional to the line center optical depth.  Near the base of the atmosphere, individual simulation cells can have optical depths of more than a million, so most of the time is spent integrating photons in these cells. This means that any method that can treat these zone via an analytic solution has the potential to greatly accelerate the calculation.

There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations.  The simplest class of methods that is specific to resonance line transfer are core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A}.  These methods focus on Monte Carlo transfer in the wings of the line and avoid scatterings in the core. Line core photons are pushed to the wings by biasing the atomic velocity distribution to high velocities so the opacity is lower and escape is more rapid.

A second class of methods are hybrid diffusion methods \citep{2001JCoPh.172..543G}, such as discrete diffusion Monte Carlo \citep{2007JCoPh.222..485D}. This involves designating sufficiently optically thick simulations cells for computation via a finite difference representation of the radiation transfer equation. This method is particularly attractive for gray opacity applications where a single diffusion equation can be computed.  For the resonant scattering problem, where the opacity is a strong function of frequency, implementation of this method is more complex and computationally demanding as it requires the solution of many frequency groups, but can lead to a considerable speed up \citep{2018MNRAS.479.2065S}.

@@

\todo{FOR REVIEW: a summative few sentences connecting the MC methods to what we're doing in this paper:}

\proposed{In this work, we present an analytic solution to the steady-state problem of \lya resonant scattering in spherical geometry and a semi-analytic solution to time-dependent \lya diffusion through the same medium. These will be shown to accurately represent the photon distributions in frequency and time-to-escape at the surface of a spherical volume of gas, highlighting their potential for use as a Monte Carlo accelerant in magnetohydrodynamic calculations of phenomena involving \lya resonance-line radiation.}
